Parallel Adaptive Importance Sampling

### 2017-01-24 PAIS.tex
Simon final changes
Small edits, typos, throughout. Couple of smallish additions.

This problem has been tackled in a variety of different ways. One
approach is to constuct increasingly complex MCMC methods which are
able to use the structure of the posterior to make more intelligent
proposals, leading to more thorough exploration of the posterior with
fewer iterations. For example, the Hamiltonian or Hybrid Monte Carlo
(HMC) algorithm uses gradient information and simplectic integrators
in order to make very large moves in state with relatively high
acceptance probability\cite{sexton1992hamiltonian}. Non-reversible
methods are also becoming quite popular as they can improve
mixing\cite{bierkens2015non}. Riemann
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.

Since the clock speed of an individual processor is no longer
following Moore's law\cite{moore1998cramming}, improvements in
computational power are largely coming from the parallelisation of
multiple cores. As such, the area of parallel MCMC methods is becoming
increasingly of interest. One class of parallel MCMC method uses
multiple proposals, with only one of these proposals being
accepted. Examples of this approach include multiple try
MCMC\cite{liu2000multiple} and ensemble MCMC\cite{neal2011mcmc}. In
\cite{calderhead2014general}, a general construction for the
parallelisation of MCMC methods was presented, which demonstrated
speed ups of up to two orders of magnitude when compared with serial
methods.

@@

\subsection{Particle filters and resamplers}\label{sec:filters}

@@

Importance sampling can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen, from which we
can draw samples. Each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use an ensemble of states to construct a proposal
distribution which will be as close as possible to the target density. If this
ensemble is large enough, the distribution of states will be representative
of the target density.

The proposal distribution could be constructed in many different ways,
but we choose to use a mixture distribution, made up of a sum of MCMC
proposal kernels. This corresponds to replacing the Dirac distribution
in the right-hand side of Equation~\ref{eqn:resampler} with a chosen
distribution e.g. Gaussian. Once the proposal is constructed, we can
sample a new ensemble from the proposal distribution, and each is
assigned a weight given by the ratio of the target density and the
proposal mixture density. Assuming that our proposal distribution is a
good one, then the variance of the weights will be small, and we will
have many useful samples. Finally, we need to create a set of evenly
weighted samples which best represent this set of weighted samples.
This is achieved by implementing a resampling algorithm. Initially we
will use the ETPF algorithm\cite{reich2013nonparametric}, although we
will suggest an alternative strategy in Section \ref{sec:AMR}. The
resampling algorithm gives us a set of evenly weighted samples which
represents the target distribution well, which we can use to iterate
the process again. The algorithm is summarised in Algorithm
\ref{alg:PAIS}.

@@

In Section \ref{Sec:Num}, we will demonstrate how the algorithm
performs, primarily using random walk (RW) proposals. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
ensemble members can improve on the original MH algorithm and lead to
tolerances being achieved in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set this could lead to a
large saving of computational cost. We have observed that
using more complex (and hence more expensive) kernels $\nu$, does not
significantly improve the speed of convergence of the algorithm for
posteriors which are easy to approximate\cite{Paul}.

@@

The effective sample size of a weighted sample is defined in the
following way:
\[
	\neff = \frac{\left(\sum_{i=1}^M \! w_i\right)^2}{\sum_{i=1}^M \! w_i^2} \approx \frac{M\E(w)^2}{\E(w^2)} = M\left(1 - \frac{\mbox{var}(w)}{\mathbb{E}(w^2)}\right).
\]
The second two expressions are true when $M\rightarrow\infty$. From
the last expression, if the variance of the weights is zero then
$\neff = M$; this is our ideal scenario. Maximising the effective
sample size is equivalent to minimising the variance of the weights.

The statistic $\neff$ is easier to deal with than the variance of the
weights, as it takes values on $[1, M]$ while the variance of the
weights takes values on $\mathbb{R}_\geq$. Since the variance can vary
over many orders of magnitude bringing numerical issues, the effective
sample size is more desirable.

@@

\begin{figure}[htb]
\centering
\subfigure[Contours showing optimal ranges of the scaling parameter.]{\includegraphics[width=0.47\textwidth]{"figures/ESSvM_contour"}}
\subfigure[The highest effective sample size achieved for each ensemble size.]{\includegraphics[width=0.47\textwidth]{"figures/ESSvM"}}
\caption{The behaviour of the effective sample size as the ensemble size increases. This analysis is of the example in Section~\ref{sec:problem 1} using the PAIS-RW algorithm.}
\label{fig:neff-M}
\end{figure}

Figure~\ref{fig:neff-M} shows how the effective sample size used in
PAIS is affected by the ensemble size. We see from subfigure (a) that
as the ensemble size increases, the optimal scaling parameter
decreases. This is expected since the scaling parameter must be
reduced if we are to fit more kernels into the same target. We also
see that the algorithm becomes less sensitive to changes in the
scaling parameter as the ensemble size increases. Subfigure (b) shows
that as the ensemble size increases, the algorithm becomes more
efficient.

@@

As outlined in~\cite{martino2015adaptive} consistency of population AIS algorithms can be considered in two different cases. In the first case we fix the number of iterations $N < \infty$, but allow the population size to grow to inifinity $M\rightarrow\infty$. In the second case we hold the population size fixed $M<\infty$, and allow infinitely many iterations $N\rightarrow\infty$.

@@

Consider the case of a linear observation operator which maps a one
dimensional parameter onto the observale space, $\G\colon x \mapsto
x$. We assign centred priors to the parameter $x$, as well as to the
observational noise. We choose the true value of the parameter to be
$x_\text{ref} = 4$, then following Equations~\eqref{eqn:obs} and
\eqref{eqn:like},
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x
	 - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right).
\end{equation}

@@

Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and
by the end of the simulation window is matching its performance. The
ARWMH algorithm does not perform quite as well, this is possibly because
it is optimising a cost functional which is not smooth.

@@

In this example we look at a bimodal posterior distribution which has
a high energy barrier between the two modes. In the MH algorithm, the
expected number of iterations for a switch between modes is a large
proportion of the total number of iterations. This posterior is in
Figure~\ref{fig:BM2_posterior}.

@@

In particular we look at a data set where we assume that there are two
subpopulations within the overall population. Since both
subpopulations will be approximated by Gaussian distributions we have
five parameters which we need to be estimated, two means $\mu_{1,2}$,
two variances $\sigma^2_{1,2}$, and the probability, $p$, that an
individual observation belongs to the first subpopulation. We have 100
data points, $D_i$, which we assume to be distributed according to
\[
	D_i \sim p\mathcal{N}(\mu_1, \sigma^2_1) + (1-p)\mathcal{N}
	(\mu_2, \sigma^2_2), \quad i = 1,\dots,100,
\]
where $p \in [0, 1]$, $\mu_{1,2} \in \mathbb{R}$ and $\sigma^2_{1,2}
\in \mathbb{R}^+$. Due to the domains of these parameters and also
some prior knowledge, we assign the priors
\[
	p \sim \text{Beta}(1,1), \quad \mu_{1,2} \sim \mathcal{N}(0, 4)
	 \quad \text{and} \quad \sigma^2_{1,2} \sim \text{Gamma}
	(\alpha=2, \beta=1).
\]
If we collect these parameters in the vector $\theta = (p, \mu_1,
\sigma^2_1, \mu_2, \sigma^2_2)^\top$, the resulting posterior
distribution is
\begin{equation}\label{eq:mixture_posterior}
	\pi(\theta|D) \propto \prod\limits_{i=1}^{100} (p\mathcal{N}
	(D_i;\mu_1, \sigma^2_1) + (1-p)\mathcal{N}(D_i;\mu_2, \sigma^2_2))
	\prod\limits_{i=1}^5 \pi_0^i(\theta_i),
\end{equation}
where $\pi_0^i(\cdot)$ is the prior density function corresponding to $\theta_i$.

@@

We have no analytic form for the posterior distribution for this
model, and MH cannot reliably sample from this type of bimodal
distribution. We do however know that the probability density should
be evenly divided between the two modes. This is due to the symmetric
prior for $p$, and and the same priors being assigned to $\mu_1$ and
$\mu_2$, and also to $\sigma^2_1$ and $\sigma^2_2$. To decide which
mode a sample belongs to we define a plane which bisects the posterior
so that each point on this plane lies exactly halfway between the two
true solutions to the inverse problem i.e. the value of $\theta$ used
to generate the data, and also the $\theta$ obtained by a relabelling
of the parameters. Now that we can assign a sample to a particular
mode, we can calculate the density in each mode by summing the weights
associated to all samples in that mode,
\[
	\bar{w}_k = \sum\limits_{i=1}^N w_i\text{I}(X_i \in \text{Mode $k$}), \quad k = 1, 2,
\]
and the relative error in the amount of density in each mode is then
\begin{equation}\label{eqn:mode_prop}
	w_\text{error} = 2\left|\frac{\bar{w}_1}{\bar{w}_1+\bar{w}_2} - \frac{1}{2}\right|.
\end{equation}

@@

Another strength of the PAIS is that it can also be used with any MCMC
proposal. There are a growing number of increasing sophisticated MCMC
algorithms (HMC, Riemann manifold MCMC etc) which could be
incorporated into this framework, leading to even more efficient
algorithms, and this is another opportunity for future work.

>>>>

This problem has been tackled in a variety of different ways. One
approach is to construct increasingly complex MCMC methods which are
able to use the structure of the posterior to make more intelligent
proposals, leading to more thorough exploration of the posterior with
fewer iterations. For example, the Hamiltonian or Hybrid Monte Carlo
(HMC) algorithm uses gradient information and symplectic integrators
in order to make very large moves in state with relatively high
acceptance probability\cite{sexton1992hamiltonian}. Non-reversible
methods are also becoming quite popular as they can improve
mixing\cite{bierkens2015non}. Riemann
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.

Since the clock speed of an individual processor is no longer
following Moore's law\cite{moore1998cramming}, improvements in
computational power are largely coming from the parallelisation of
multiple cores. As such, the area of parallel MCMC methods is becoming
increasingly of interest. One class of parallel MCMC method uses
multiple proposals, with only one of these proposals being
accepted. Examples of this approach include multiple try
MCMC\cite{liu2000multiple} and ensemble MCMC\cite{neal2011mcmc}. In
\cite{calderhead2014general}, a general construction for the
parallelisation of MCMC methods was presented, which demonstrated
speed ups of up to two orders of magnitude when compared with serial
methods on a single core.

@@

\subsection{Particle filters and resamplers}\label{sec:filters}
In this subsection we briefly review particle filters, since the
development of the resampler that we incorporate into the PAIS is
motivated by this area.

@@

Importance sampling can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen, from which we
can draw samples. Each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use an ensemble of states to construct a proposal
distribution which will be as close as possible to the target density. If this
ensemble is large enough, the distribution of states will be
approximately representative
of the target density.

The proposal distribution could be constructed in many different ways,
but we choose to use a mixture distribution, made up of a sum of MCMC
proposal kernels. This corresponds to replacing the Dirac distribution
in the right-hand side of Equation~\ref{eqn:resampler} with a chosen
distribution e.g. Gaussian. Once the proposal is constructed, we can
sample a new ensemble from the proposal distribution, and each is
assigned a weight given by the ratio of the target density and the
proposal mixture density. Assuming that our proposal distribution is a
good one, then the variance of the weights will be small, and we will
have many useful samples. Finally, we need to create a set of evenly
weighted samples which best represent this set of weighted samples.
This is achieved by implementing a resampling algorithm. These samples
are not stored in order to characterise the posterior density, since
the resampling process inevitably adds some error/bias. They are
simply needed in order to inform the mixture distribution for the next
iteration of the algorithm.

Initially we
will use the ETPF algorithm\cite{reich2013nonparametric}, although we
will suggest an alternative strategy in Section \ref{sec:AMR}. The
resampling algorithm gives us a set of evenly weighted samples which
represents the target distribution well, which we can use to iterate
the process again. The algorithm is summarised in Algorithm
\ref{alg:PAIS}.

@@

In Section \ref{Sec:Num}, we will demonstrate how the algorithm
performs, primarily using random walk (RW) proposals. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
ensemble members can improve on the original MH algorithm and lead to
tolerances being achieved in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set this could lead to a
large saving of computational cost. We have observed that
using more complex (and hence more expensive) kernels $\nu$, does not
significantly improve the speed of convergence of the algorithm for
posteriors that we have considered\cite{Paul}.

@@

The effective sample size of a weighted sample is defined in the
following way:
\[
	\neff = \frac{\left(\sum_{i=1}^M \! w_i\right)^2}{\sum_{i=1}^M \! w_i^2} \approx \frac{M\E(w)^2}{\E(w^2)} = M\left(1 - \frac{\mbox{var}(w)}{\mathbb{E}(w^2)}\right).
\]
The second two expressions are true when $M\rightarrow\infty$. From
the last expression, if the variance of the weights is zero then
$\neff = M$; this is our ideal scenario. In the limit
$M\rightarrow\infty$, maximising the effective
sample size is equivalent to minimising the variance of the weights.

The statistic $\neff$ is easier to deal with than the variance of the
weights, as it takes values on $[1, M]$ while the variance of the
weights takes values on $\mathbb{R}_\geq$. Moreover the variance of
the weights can vary
over many orders of magnitude causing numerical issues, so that the effective
sample size is more desirable as an indicator of optimality.

@@

\begin{figure}[htb]
\centering
\subfigure[Contours showing optimal ranges of the scaling parameter.]{\includegraphics[width=0.47\textwidth]{"figures/ESSvM_contour"}}
\subfigure[The highest effective sample size achieved for each ensemble size.]{\includegraphics[width=0.47\textwidth]{"figures/ESSvM"}}
\caption{The behaviour of the effective sample size as the ensemble
  size increases, considering the example in Section~\ref{sec:problem 1} using the PAIS-RW algorithm.}
\label{fig:neff-M}
\end{figure}

Figure~\ref{fig:neff-M} shows how the effective sample size used in
PAIS is affected by the ensemble size. We see from subfigure (a) that
as the ensemble size increases, the optimal scaling parameter
decreases. This is expected since the larger ensemble allows for
finer resolution in the proposal distribution approximation of the
posterior distribution. We also
see that the algorithm becomes less sensitive to changes in the
scaling parameter as the ensemble size increases. Subfigure (b) shows
that as the ensemble size increases, the algorithm becomes more
efficient.

@@

As outlined in~\cite{martino2015adaptive} consistency of population AIS algorithms can be considered in two different cases. In the first case we fix the number of iterations $N < \infty$, but allow the population size to grow to infinity $M\rightarrow\infty$. In the second case we hold the population size fixed $M<\infty$, and allow infinitely many iterations $N\rightarrow\infty$.

@@

Consider the case of a linear observation operator which maps a one
dimensional parameter onto the observable space, $\G\colon x \mapsto
x$. We assign centred priors to the parameter $x$, as well as to the
observational noise. We choose the true value of the parameter to be
$x_\text{ref} = 4$, then following Equations~\eqref{eqn:obs} and
\eqref{eqn:like},
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x
	 - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right).
\end{equation}

@@

Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and
by the end of the simulation window is matching its performance. The
ARWMH algorithm does not perform quite as well, this is possibly because
it is optimising a cost functional which is not smooth due to an
increased sensitivity to Monte Carlo error.

@@

In this example we look at a bimodal posterior distribution which has
a high energy barrier between the two modes. In the MH algorithm, the
expected number of iterations for a switch between modes is a large
proportion of the total number of iterations. This posterior is presented in
Figure~\ref{fig:BM2_posterior}.

@@

In particular we look at a data set where we assume that there are two
subpopulations within the overall population. Since both
subpopulations will be approximated by Gaussian distributions we have
five parameters which we need to be estimated, two means $\{\mu_{1}, \mu_{2}\}$,
two variances $\{\sigma^2_{1}, \sigma^2_{2}\}$, and the probability, $p$, that an
individual observation belongs to the first subpopulation. We have 100
data points, $D_i$, which we assume to be distributed according to
\[
	D_i \sim p\mathcal{N}(\mu_1, \sigma^2_1) + (1-p)\mathcal{N}
	(\mu_2, \sigma^2_2), \quad i = 1,\dots,100,
\]
where $p \in [0, 1]$, $\mu_{1},\mu_2 \in \mathbb{R}$ and $\sigma^2_{1},\sigma^2_2
\in \mathbb{R}^+$. Due to the domains of these parameters and also
some prior knowledge, we assign the priors
\[
	p \sim \text{Beta}(1,1), \quad \mu_{1,2} \sim \mathcal{N}(0, 4)
	 \quad \text{and} \quad \sigma^2_{1,2} \sim \text{Gamma}
	(\alpha=2, \beta=1).
\]
If we collect these parameters in the vector $\theta = (p, \mu_1,
\sigma^2_1, \mu_2, \sigma^2_2)^\top$, the resulting posterior
distribution is
\begin{equation}\label{eq:mixture_posterior}
	\pi(\theta|D) \propto \prod\limits_{i=1}^{100} (p\mathcal{N}
	(D_i;\mu_1, \sigma^2_1) + (1-p)\mathcal{N}(D_i;\mu_2, \sigma^2_2))
	\prod\limits_{i=1}^5 \pi_0^i(\theta_i),
\end{equation}
where $\pi_0^i(\cdot)$ is the prior density function corresponding to $\theta_i$.

@@

We have no analytic form for the posterior distribution for this
model, and MH cannot reliably sample from this type of bimodal
distribution. We do however know that the probability density should
be evenly divided between the two modes. This is due to the symmetric
prior for $p$, and and the same priors being assigned to $\mu_1$ and
$\mu_2$, and also to $\sigma^2_1$ and $\sigma^2_2$. To decide which
mode a sample belongs to we define a plane which bisects the posterior
so that each point on this plane lies exactly halfway between the two
true solutions to the inverse problem i.e. the value of $\theta$ used
to generate the data, and also the $\theta$ obtained by a relabelling
of the parameters. Now that we can assign a sample to a particular
mode, we can calculate the density in each mode by summing the weights
associated to all samples in that mode,
\[
	\bar{w}_k = \sum\limits_{i=1}^N w_iI_{X_i \in \text{Mode $k$}}, \quad k = 1, 2,
\]
and the relative error in the amount of density in each mode is then
\begin{equation}\label{eqn:mode_prop}
	w_\text{error} = 2\left|\frac{\bar{w}_1}{\bar{w}_1+\bar{w}_2} - \frac{1}{2}\right|.
\end{equation}

@@

Another strength of the PAIS is that it can also be used with any MCMC
proposal. There are a growing number of increasing sophisticated MCMC
algorithms (non-reversible Langevin/HMC proposals, Riemann manifold MCMC etc) which could be
incorporated into this framework, leading to even more efficient
algorithms, and this is another opportunity for future work.
